---
title: "Bilobate Presence, no pseudorandom"
author: "Steve Fick"
date: "September 10, 2015"
output:
  html_document: default
  pdf_document: null
  toc: yes
---
  
  **Response: Bilobate Presence, no pseudorandom**

### Some Definitions
  
* **Phytolith** : Phytoliths are microscopic, weathering-resistant particles of
silica formed in certain plant taxa that are released from decomposing
litter and accumulate over time in the soil

* **rondel_cells** : Exotic grasses only produce rondel-type phytoliths  

* **bilobate cells** : only produced by proposed native bunch grasses (not produced by natives)  

* **short cells ** : specialized silica accumulating cells found only in
grasses. These strongly correlate to total phytolith mass (in evett 2013), indicating that most of the phytoliths produced were from grass cells.    

* **total phytolith ** : Total mass of phytoliths.  Evett et al. (2012) proposed distinguishing sites with substantial long-term grass
cover from other sites by using a ??? 0.3% soil phytolith weight threshold

* **bilobate ratio ** : Percentage dominated by stipa and /or Danthonia

* ***Stipa pulchra*** : Proposed dominant grass species in Central Valley  

* ***Danthonia californica*** : proposed dominant grass in Coastal areas  


```{r, include = FALSE, cache=FALSE}
  
  # some required packages
  library(maps)
  library(knitr)
  library(raster)
  library(dismo)
  library(arm)
  library(MASS)
  library(glmnet)
  library(randomForest)
  library(rpart)
  library(spdep)
  library(xtable)
  require(rpart.plot)
```

```{r setup, include = FALSE, cache=TRUE}
 
  setwd('C:/projects/phytolith/')
  
  ### Basic Data Processing
  
  p <- read.csv('data/dat.csv', stringsAsFactors = F)
  
  # remove records with missing spatial information
  p <- p[-which(is.na(p$lon)),]
  
  # reverse longitude
  p$lon <- -1*p$lon
  
  # Grass threshold according to Evett and Bartolome
  p$grass <- as.factor(p$phyto_content >= .3)
  
  # Had bilobate cells (only produced by PNG)
  p$PNG <- as.numeric(p$bilobate_cells > 0)
  
  # fix bilob_ratio
  p$bilob_ratio <- gsub('[*]','', p$bilob_ratio)
  p$bilob_ratio <- as.numeric(p$bilob_ratio)
  
  # create a 'R::spatial' version of p
  sp <- p
  coordinates(sp) <- ~ lon + lat
  
  # create a distance matrix between points to consider spatial autocorrelation in models
  dd <- as.matrix(as.dist(pointDistance(sp, lonlat = TRUE, allpairs=TRUE)))
  dd.inv <- 1/dd
  diag(dd.inv) <- 0
  
  
  ### Load Covariates
  
  b <- brick('C:/projects/phytolith/data/predictors.grd')
  mb <- brick('C:/projects/phytolith/data/mpredictors.grd')
  ca <- shapefile('C:/Users/steve/Downloads/GEO2015/13/counties_2000.shp')
  
  # Extract covariates from location
  X <- as.data.frame(extract(b, sp))
  
  # fix points which fall outside of slope raster
  i <- which(is.na(X$slope))
  X$slope[i] <- 0
  X$northness[i] <- X$eastness[i] <-pi/2 
  
  ### Some further covariate processing
  
  # remove veg and hillshade columns
  XX <- X[ , -which(names(X) %in% c('veg', 'hillshade'))]
  
  #only use pre-industrial bioclim variables
  XX <- XX[ , -grep('bio[1-9]',names(XX))]
  XX <- XX[ , -grep('bioclim_6',names(XX))]
  XX <- XX[ , -grep('bioclim_4',names(XX))]
  
  #exclude some soil covariates which likely vary on short timescales / be highly correlated with bioclim variables
  XX <- XX[ , -which(names(XX) %in% c('carbon','ph','cec'))]
  
  ### Load Functions
  
  source('code/functions.R')
  
  #################################################
  #################################################
  
  # Total phytolith content (amount of grasses)
  
  
  # identify response variable
  XX$y <- as.factor(p[, 'bilobate_cells'] > 0)
  depvarname <- 'Bilobate Presence'
  
  
	add.Psuedoabsence <- FALSE

	if( add.Psuedoabsence){

	  XX2 <- as.data.frame(sampleRandom(b, 1000, ext = extent(sp)))
	  XX2$y <- as.factor(FALSE)
	  XX2 <- XX2[ , -which(names(XX2) %in% c('veg', 'hillshade'))]
	  
	  #only use 400AD bioclim variables
	  XX <- XX[ , -grep('bio[1-9]',names(XX))]
	  XX2 <- XX2[ , -grep('bioclim_6',names(XX2))]
	  XX2 <- XX2[ , -grep('bioclim_4',names(XX2))]

	  #exclude some soil covariates which likely vary on short timescales
	  XX2 <- XX2[ , -which(names(XX2) %in% c('carbon','ph','cec'))]


	  XX <- rbind(XX2,XX)
	}


  
  f <- y ~ .  
  xn <- which(names (XX) != 'y')
  yn <- which(names(XX) == 'y')
  
 
  # fit the ensemble model
  ens <- ensemble.fit.bin(f, XX, XX[,xn],XX[,yn])
  
  # Make a prediction
  p.ens <- ensemble.pred.bin(ens, XX, XX[,xn], XX[,yn], wts = wts)
  
```

----

# Model Validation


----

# Predictions


### predictions

```{r predictions, include = FALSE, cache = TRUE}
	#random forest prediction
	p.rf <- predict(mb, ens$RF, type = 'prob', progress = 'text')

	#OLS prediction
	p.ols <- predict(mb, ens$OLS, type = 'response')

	#CT prediction
	p.ct <- predict(mb, ens$CT)

	#ensemble prediction
	m.ens <- p.ols * .33 + (1-p.ct) * .33 + (1-p.rf) * .33 

```

### Random Forest
```{r echo = FALSE}

### Surface Visualization

  #random forest prediction
	plot(1-p.rf,  main = 'Predicted N. pulchra/danthonia presence \n(random forest)' ); plot(ca, add=TRUE, border = rgb(0,0,0,.1))
```
       
       
### CART 
```{r echo = FALSE}
plot(1-p.ct,  main = 'Predicted N. pulchra/danthonia presence \n(CART)' ); plot(ca, add=TRUE, border = rgb(0,0,0,.1))
```

### OLS
```{r echo = FALSE}
plot(p.ols, main = 'Predicted N. pulchra/danthonia presence \n(OLS regression)' ); plot(ca, add=TRUE, border = rgb(0,0,0,.1))
```

### Ensemble
```{r echo = FALSE}

	plot(m.ens, col = gray(100:1/100), main = 'Predicted N. pulchra/danthonia presence  \n(Ensemble Model)' ); plot(ca, add=TRUE, border = rgb(0,0,0,.1))

	plot(m.ens, col = terrain.colors(5), main = 'Predicted N. pulchra/danthonia presence  \n(Ensemble Model)' ); plot(ca, add=TRUE, border = rgb(0,0,0,.1))

```

----

# Variable Importance

### Random Forest
```{r echo = FALSE}	
	vn <- names(ens$RF$forest$xlevels)	
	vn <- changeVarNames(vn)
	#varImpPlot(ens$RF , n.var = 10, labels = rev(vn[order(importance(ens$RF), decreasing = TRUE)][1:10]), main = paste0('Random Forest Variable Importance for', depvarname))
	varImpPlot(ens$RF ,  labels = rev(vn[order(importance(ens$RF), decreasing = TRUE)]), main = 'Random Forest Variable Importance (N. pulchra presence)')

```

### Cart
```{r echo = FALSE}	
	X2 <- XX
	names(X2) <-gsub('bioclim_400AD.', 'bio', names(X2))
	names(X2) <-gsub("bio_preind.", 'bio', names(X2))
	names(X2) <-gsub('bioclim_6kbp.', 'bio', names(X2))
	names(X2) <- ifelse(!names(X2) %in% names(codex),names(X2), codex[names(X2)])
	CT2 <- update(ens$CT, data = X2)

	prp(CT2, varlen = 0, uniform = FALSE, main = paste0('CART results for ', depvarname))
```

### OLS
```{r echo = FALSE}
	XX.scaled <- as.data.frame(scale(XX[,-which(names(XX)=='y')]))
	XX.scaled$y <- XX$y
	OLS.s <- update(ens$OLS, data = XX.scaled)
	vn2 <- names(OLS.s$coefficients)
  vn2 <- gsub('bioclim_400AD.', 'bio', vn2)
  vn2 <- gsub("bio_preind.", 'bio', vn2)
  vn2 <- gsub('bioclim_6kbp.', 'bio', vn2)

  vn2 <- ifelse(!vn2 %in% names(codex),vn2, codex[vn2])

  vn2[ which(vn2 == 'clay')] <- "clay content"
  vn2[ which(vn2 == 'bulk')] <- "Bulk density"
  vn2[ which(vn2 == 'alt')] <- "elevation"

	coefplot(OLS.s, mar=c(1,15,5.1,2), varnames = vn2, main = paste0('Scaled OLS coefficients for', depvarname))
```

----
# changes due to climate

```{r echo = FALSE}



```


----

```{r saveit}
save(p, p.ct, ens, p.rf, p.ols, p.lasso, p.ridge, p.ens, file = paste0(depvarname,'.Rmd'))


```
